---
title: "Linear Model with R: Chapter 6"
author: Han Chenyue
editor: source
format:
  html:
    code-fold: true
    code-summary: "Show the code"
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  # echo = FALSE,
  digits = 4,
  scipen = 2
)
```

# Chapter 6

## Exercises

### Q1

```{r}
data(teengamb, package = "faraway")

# ============================================================
# STEP 1: Create an indicator matrix of all possible
# combinations of including/excluding 3 predictors.
#
# expand.grid() generates every combination of the values
# we give it. Here, each predictor can be either:
#   0 = "do NOT include this predictor"
#   1 = "DO include this predictor"
#
# Since we have 3 optional predictors (status, income, verbal),
# there are 2 × 2 × 2 = 8 possible combinations.
# ============================================================
indicator_matrix <- expand.grid(0:1, 0:1, 0:1)

# Let's name the columns so we know which predictor each refers to
colnames(indicator_matrix) <- c("status", "income", "verbal")

# Print it to see all 8 combinations
# The output looks like this:
#   status income verbal
# 1      0      0      0   ← No extra predictors (just sex)
# 2      1      0      0   ← Include status only
# 3      0      1      0   ← Include income only
# 4      1      1      0   ← Include status and income
# 5      0      0      1   ← Include verbal only
# 6      1      0      1   ← Include status and verbal
# 7      0      1      1   ← Include income and verbal
# 8      1      1      1   ← Include all three
indicator_matrix
```

```{r}
# ============================================================
# STEP 2: Set up storage for our results
# ============================================================

# This is the base model formula that every model will include.
# Sex is ALWAYS a predictor — we want to see how its effect
# changes depending on what other predictors we add.
base_model_formula <- "gamble ~ sex"

# How many models are we going to fit?
# One for each row of the indicator matrix = 8
number_of_models <- nrow(indicator_matrix)

# Create an empty character vector to store the formula string
# for each model. We use rep() to repeat an empty string ""
# eight times (once per model).
#
# rep(x, times) — repeats the value x a total of 'times' times
# 'times' and 'length.out' do the same thing here;
# you only need one of them, not both!
formula_strings <- rep(x = "", times = number_of_models)

# Create an empty numeric vector to store the coefficient
# (slope) of sex for each model. We fill it with NA for now.
sex_coefficients <- rep(x = NA, times = number_of_models)

# Create an empty numeric vector to store the p-value
# of sex for each model. We fill it with NA for now.
sex_p_values <- rep(x = NA, times = number_of_models)

# ============================================================
# STEP 3: Loop through all 8 model combinations
# ============================================================

for (i in 1:number_of_models) {
  # ----------------------------------------------------------
  # STEP 3a: Build the formula string for model i
  # ----------------------------------------------------------

  # Grab the i-th row of the indicator matrix.
  # This tells us which optional predictors to include.
  current_row <- indicator_matrix[i, ]

  # Example: if i = 4, current_row is (1, 1, 0),
  # meaning: include status (1), include income (1),
  # exclude verbal (0).

  # Convert the 0s and 1s to TRUE/FALSE
  # 0 becomes FALSE, 1 becomes TRUE
  include_flags <- as.logical(current_row)

  # Example: c(1, 1, 0) becomes c(TRUE, TRUE, FALSE)

  # Check: does this row have ALL zeros?
  # (i.e., no extra predictors at all?)
  no_extra_predictors <- !any(include_flags)
  # any() returns TRUE if at least one element is TRUE
  # !any() returns TRUE only if ALL elements are FALSE

  if (no_extra_predictors) {
    # If no extra predictors, use the base model: gamble ~ sex
    current_formula_string <- base_model_formula
  } else {
    # Define the full list of optional predictor names
    optional_predictor_names <- c("status", "income", "verbal")

    # Use the TRUE/FALSE flags to select which predictors
    # to include. This is called "logical subsetting."
    #
    # Example: c("status","income","verbal")[c(TRUE,TRUE,FALSE)]
    #   → returns c("status", "income")
    selected_predictor_names <- optional_predictor_names[include_flags]

    # Collapse the selected predictor names into a single
    # string separated by "+"
    #
    # Example: c("status", "income") becomes "status+income"
    extra_predictors_string <- paste(selected_predictor_names, collapse = "+")

    # Now glue the base formula together with the extra
    # predictors, separated by "+"
    #
    # Example: paste("gamble ~ sex", "status+income", sep = "+")
    #   → "gamble ~ sex+status+income"
    current_formula_string <- paste(
      base_model_formula,
      extra_predictors_string,
      sep = "+"
    )
  }

  # ----------------------------------------------------------
  # STEP 3b: Convert string to formula and fit the model
  # ----------------------------------------------------------

  # R's lm() function needs a formula object, not a plain string.
  # as.formula() converts a string like "gamble ~ sex+income"
  # into a proper R formula object.
  current_formula_object <- as.formula(current_formula_string)

  # Fit the linear regression model using lm()
  fitted_model <- lm(formula = current_formula_object, data = teengamb)

  # Get the detailed summary of the model, which includes
  # coefficient estimates, standard errors, t-values, and p-values
  model_summary <- summary(fitted_model)

  # ----------------------------------------------------------
  # STEP 3c: Extract the sex coefficient and its p-value
  # ----------------------------------------------------------

  # model_summary$coefficients is a matrix that looks like:
  #
  #              Estimate  Std.Error  t.value   Pr(>|t|)
  # (Intercept)   ...       ...        ...       ...
  # sex           ...       ...        ...       ...
  # income        ...       ...        ...       ...
  # ...
  #
  # Row 1 = Intercept
  # Row 2 = sex (because sex is the FIRST predictor listed
  #              in every formula, right after the intercept)
  # Column 1 = Estimate (the coefficient value)
  # Column 4 = Pr(>|t|) (the p-value)

  coefficient_table <- model_summary$coefficients

  # The sex coefficient estimate is in row 2, column 1
  sex_estimate <- coefficient_table[2, 1]

  # The sex p-value is in row 2, column 4
  sex_pvalue <- coefficient_table[2, 4]

  # ----------------------------------------------------------
  # STEP 3d: Store the results for model i
  # ----------------------------------------------------------

  formula_strings[i] <- current_formula_string
  sex_coefficients[i] <- sex_estimate
  sex_p_values[i] <- sex_pvalue

  # Optional: print progress so we can watch it work
  cat(
    "Model",
    i,
    ":",
    current_formula_string,
    " | sex coef =",
    round(sex_estimate, 2),
    " | p-value =",
    round(sex_pvalue, 4),
    "\n"
  )
}

# ============================================================
# STEP 4: Assemble and display the results in a nice table
# ============================================================

results_table <- data.frame(
  model = formula_strings, # The formula used
  sex_coefficient = sex_coefficients, # The estimate for sex
  sex_p_value = sex_p_values # The p-value for sex
)

# Print the final table
results_table
```

Across all eight models, the sex coefficient is always negative (*females gamble less*) and always statistically significant (*small p-value*). This means the effect of sex is stable and robust, and it doesn't matter which combination of the other predictors (*status, income, verbal*) we include or exclude.

### Q2

```{r}
data(odor, package = "faraway")

lm_q2_list <- c(
  "odor ~ temp",
  "odor ~ temp + gas",
  "odor ~ temp + pack",
  "odor ~ temp + gas + pack"
)

linear_models_matrix <- NULL

for (i in seq_along(lm_q2_list)) {
  lm <- lm(as.formula(lm_q2_list[i]), data = odor)
  linear_models_matrix <- rbind(
    linear_models_matrix,
    summary(lm)$coef["temp", ]
  )
}

row.names(linear_models_matrix) = lm_q2_list
linear_models_matrix
```

- The coefficient estimate for `temp` is exactly `r linear_models_matrix[1, "Estimate"]` in all four models. No matter which combination of `gas` and `pack` we include, the estimated effect of temperature on odor does not change at all.
- The **standard error decreases** as we add more predictors (`r round(linear_models_matrix[1, "Std. Error"], 2)` → `r round(linear_models_matrix[2, "Std. Error"], 2)` → `r round(linear_models_matrix[3, "Std. Error"], 2)` → `r round(linear_models_matrix[4, "Std. Error"], 2)`)
- The **t-statistic increases in magnitude** (`r round(linear_models_matrix[1, "t value"], 3)` → `r round(linear_models_matrix[2, "t value"], 3)` → `r round(linear_models_matrix[3, "t value"], 3)` → `r round(linear_models_matrix[4, "t value"], 3)`)
- The **p-value decreases** (`r round(linear_models_matrix[1, "Pr(>|t|)"], 3)` → `r round(linear_models_matrix[2, "Pr(>|t|)"], 3)` → `r round(linear_models_matrix[3, "Pr(>|t|)"], 3)` → `r round(linear_models_matrix[4, "Pr(>|t|)"], 3)`)
- However, `temp` remains **non-significant in all four models** (*all p-values > 0.05*).
- The coefficient for `temp` is perfectly stable because the predictors in the `odor` dataset are **orthogonal** (uncorrelated) — this is a **designed experiment** where the factor levels were deliberately chosen to be balanced and independent. When predictors are orthogonal, adding or removing other predictors **cannot change** the coefficient of `temp`, because there is no shared variance between the predictors that would cause confounding.
- The standard error decreases because adding useful predictors (`gas`, `pack`) **reduces the residual variance** (RSS). Since the standard error of a coefficient is proportional to the residual standard error ($\hat{\sigma}$), a smaller RSS → smaller $\hat{\sigma}$ → smaller standard error → larger $|t|$ → smaller p-value.
-The **full model** (`odor ~ temp + gas + pack`) is best because:
  - It has the **smallest standard error** for `temp`, giving us the most **precise** estimate of its effect.
  - Adding `gas` and `pack` reduces residual variance, improving the model's overall explanatory power.
  - In an orthogonal design, there is **no cost** from multicollinearity when adding predictors — only the benefit of reduced residual variance.
- Orthogonal designs are highly desirable in experiments precisely because of this property — the estimated effect of each factor is **completely independent** of which other factors are in the model, making the analysis clean and interpretation straightforward.

### Q3

#### q3a

```{r}
plot(gamble ~ income, data = teengamb, pch = sex)
legend("topleft", legend = c("Male", "Female"), pch = 0:1)
```

#### q3b

```{r}
lm_q3b <- lm(gamble ~ income + sex, data = teengamb)
lm_q3b_coefficients <- coef(object = lm_q3b)
plot(gamble ~ income, data = teengamb, pch = sex)
abline(a = lm_q3b_coefficients[1], b = lm_q3b_coefficients[2], lty = 1)
abline(
  a = lm_q3b_coefficients[1] + lm_q3b_coefficients[3],
  b = lm_q3b_coefficients[2],
  lty = 2
)
legend("topleft", legend = c("Male", "Female"), pch = 0:1, lty = 1:2)
```

#### q3c

```{r}
mattg = merge(
  x = teengamb[teengamb$sex == 1, c(3, 5)],
  y = teengamb[teengamb$sex == 0, c(3, 5)],
  by = "income",
  suffixes = c("female", "male")
)
mattg
```

```{r}
nrow(mattg)
```

There are `r nrow(mattg)` matched pairs. However, the `teengamb` dataset only has `r nrow(teengamb)` individuals in total. Since each pair consists of one male and one female matched on income, the `r nrow(mattg)` pairs imply `r nrow(mattg) * 2` individual slots — more than the `r nrow(teengamb)` people available. This means some individuals appear in more than one matched pair, because `merge()` performs a many-to-many join: if multiple people of one sex share the same income as someone of the other sex, all possible pairings are created.

#### q3d

```{r}
t.test(mattg$gamblemale - mattg$gamblefemale)
```

The paired t-test yields a t-statistic of `r round(t.test(mattg$gamblemale - mattg$gamblefemale)$statistic, 4)` with a p-value of `r round(t.test(mattg$gamblemale - mattg$gamblefemale)$p.value, 4)`. At the 5% significance level, the difference between male and female gambling expenditure is **not statistically significant**, though the result is borderline with a p-value of approximately 0.064.

However, a key assumption of the paired t-test is **independence between the pairs**. This assumption is clearly violated here because, as we noted above, some individuals appear in more than one matched pair due to the many-to-many nature of the `merge()`. This means the differences across pairs are **not independent** — if a particular male appears in multiple pairs, the differences involving that male will be correlated. As a result, the standard error and p-value from this test are **unreliable**, and the conclusion should be interpreted with caution.

#### q3e

```{r}
summary(lm_q3b)
```

From the model in (b), the estimated coefficient for `sex` is `r round(summary(lm_q3b)$coef["sex", "Estimate"], 2)`, meaning that females gamble approximately `r abs(round(summary(lm_q3b)$coef["sex", "Estimate"], 2))` pounds less on average than males after adjusting for income. This difference is statistically significant with a p-value of `r round(summary(lm_q3b)$coef["sex", "Pr(>|t|)"], 4)`.

Compared to the paired t-test in (d), where the estimated difference was approximately `r round(mean(mattg$gamblemale - mattg$gamblefemale), 2)` pounds and was not statistically significant (p ≈ `r round(t.test(mattg$gamblemale - mattg$gamblefemale)$p.value, 4)`), the regression approach finds a **larger** and **statistically significant** difference. This discrepancy arises because the paired t-test in (d) was compromised by the violation of the independence assumption — the reuse of individuals across multiple matched pairs introduced correlated observations, inflating the standard error and weakening the test's ability to detect a true difference.

#### q3f

```{r}
plot(
  gamblemale - gamblefemale ~ income,
  data = mattg,
  ylab = "Differences in amount gambled by sex (better way to phrase this?)"
)
lm_q3f <- lm(gamblemale - gamblefemale ~ income, data = mattg)
abline(lm_q3f)
summary(lm_q3f)
```

The plot shows a clear positive trend: the difference in gambling expenditure between matched males and females **increases with income**. The fitted regression line has a slope of `r round(summary(lm_q3f)$coef["income", "Estimate"], 2)`, suggesting that for each unit increase in income, the male-female gambling gap widens by approximately `r round(summary(lm_q3f)$coef["income", "Estimate"], 2)` pounds.

This is consistent with what we observed in the initial plot from (a), where the **gap between the male and female regression lines widens** as income increases — the male line has a steeper slope than the female line, meaning higher-income males gamble disproportionately more than their female counterparts.

#### q3g

```{r}
plot(gamble ~ income, data = teengamb, pch = sex)
legend("topleft", legend = c("Male", "Female"), pch = 0:1, lty = 1:2)
lm_q3g_i = lm(gamble ~ income, data = teengamb, subset = (sex == 1))
abline(lm_q3g_i, lty = 2)
lm_q3g_ii <- lm(gamble ~ income, data = teengamb, subset = (sex == 0))
abline(lm_q3g_ii)
```

The plot reveals a striking difference in the relationship between gambling and income for males versus females. The male regression line has a steep positive slope of `r round(coef(lm_q3g_ii)[2], 2)`, indicating that male gambling expenditure increases substantially with income. In contrast, the female regression line is nearly flat with a slope of `r round(coef(lm_q3g_i)[2], 2)`, suggesting that female gambling expenditure barely changes regardless of income level.

As a result, the gap between the two lines **widens as income increases** — at low income levels, males and females gamble roughly similar amounts, but at higher income levels, males gamble far more than females. This is entirely consistent with the finding in (f), where the male-female difference in matched pairs increased with income.

Together, (f) and (g) suggest that the additive model in (b), which assumes parallel lines (i.e., a constant difference between sexes across all income levels), is **inadequate**. The non-parallel lines seen here indicate that the effect of sex on gambling **depends on income**, meaning an interaction term (`sex * income`) should be included in the model to properly capture this relationship.

### Q4

#### q4a

```{r}
data(mba, package = "faraway")
lm_q4a <- lm(happy ~ money + sex + love + work, data = mba)
summary(lm_q4a)
```

A unit increase in `love`, with the other predictors held constant, is associated with an estimated increase of `r round(summary(lm_q4a)$coefficients[4], 2)` in the response `happy`.

#### q4b

```{r}
mba$clove <- ifelse(test = mba$love <= 2, yes = 0, no = 1)
lm_q4b <- lm(happy ~ money + sex + clove + work, data = mba)
summary(lm_q4b)
```

The new predictor `clove` is a binary variable that equals 0 if `love` ≤ 2 and 1 if `love` = 3. Since `clove` only has two levels, its coefficient has a clear interpretation: holding all other predictors constant, individuals with the higher level of love (`clove` = 1) are estimated to have a `happy` score that is `r round(summary(lm_q4b)$coef["clove", "Estimate"], 2)` units higher than those with the lower level of love (`clove` = 0). This effect is highly significant (p-value equals to `r format(summary(lm_q4b)$coef["clove", "Pr(>|t|)"], digits = 3)`).

Comparing the two models, the results are broadly similar:

- Both `clove` and `love` are highly significant predictors of `happy` (p < 0.001 in both cases).
- The significance and direction of the other predictors (`money`, `sex`, `work`) remain largely unchanged.
- The $R^2$ is slightly lower in the `clove` model (`r round(summary(lm_q4b)$r.squared, 4)` vs. `r round(summary(lm_q4a)$r.squared, 4)`), indicating a modest loss in explanatory power.

The results do not differ much because most of the variation in `love` is between the values 2 and 3, which is exactly the distinction that `clove` captures. By collapsing `love` values of 1 and 2 into a single category, we lose only a small amount of information — specifically, the distinction between `love` = 1 and `love` = 2, which occurs rarely in the data. The slight decrease in $R^2$ and increase in residual standard error (`r round(summary(lm_q4b)$sigma, 3)` vs. `r round(summary(lm_q4a)$sigma, 3)`) reflect this minor information loss.

#### q4c

```{r}
lm_q4c <- lm(happy ~ clove, data = mba)
summary(lm_q4c)
```

With only `clove` as a predictor, subjects with the higher level of love (`clove` = 1) are estimated to have a `happy` score that is `r round(summary(lm_q4c)$coef["clove", "Estimate"], 2)` units higher than those with the lower level (`clove` = 0). The effect remains highly significant (p = `r format(summary(lm_q4c)$coef["clove", "Pr(>|t|)"], digits = 3)`).

Compared to the model in (b), the coefficient for `clove` is **larger** here (`r round(summary(lm_q4c)$coef["clove", "Estimate"], 2)` vs. `r round(summary(lm_q4b)$coef["clove", "Estimate"], 2)`). This is because we are no longer adjusting for the other predictors (`money`, `sex`, `work`). In the multiple regression model of (b), the effect of `clove` was estimated while **holding the other variables constant**, effectively isolating its unique contribution. Here, with `clove` as the sole predictor, its coefficient absorbs some of the effects that the other predictors would have explained — particularly those that are correlated with both `clove` and `happy`. This results in a larger but **less precise** estimate, as reflected in the higher residual standard error (`r round(summary(lm_q4c)$sigma, 3)` vs. `r round(summary(lm_q4b)$sigma, 3)`) and lower $R^2$ (`r round(summary(lm_q4c)$r.squared, 4)` vs. `r round(summary(lm_q4b)$r.squared, 4)`).

#### q4d

```{r}
plot(
  happy ~ jitter(work),
  data = mba,
  pch = clove,
  xlab = "Work",
  ylab = "Happy"
)
legend("topleft", legend = c("Low Love", "High Love"), pch = 0:1)
```

#### q4e

```{r}
xtabs(~ clove + work, data = mba)
```

This gives a maximum of $1 + 1 + 5 + 7 + 0 = 14$ one-to-one matched pairs.

#### q4f

```{r}
# Computing the mean `happy` for each combination of `work` and `clove`
means_of_work_and_clove <- aggregate(
  x = happy ~ work + clove,
  data = mba,
  FUN = mean
)
means_of_work_and_clove
```

```{r}
# Then we compute the difference in mean happy between clove = 1 and clove = 0 at each level of work
# work = 5 is excluded from this calculation because there are no clove = 0 individuals at that level
mean_differences <- means_of_work_and_clove[5:8, "happy"] -
  means_of_work_and_clove[1:4, "happy"]
mean_differences
```

```{r}
# Overall mean
mean(mean_differences)
```

This gives `r round(mean(mean_differences), 2)`, which represents a simple unweighted average of the `clove` effect across work levels.

Compare it with the following linear model on the effect of `clove` while adjusting for `work` predictor

```{r}
# Now we compare this to the linear model that estimates the effect of clove while adjusting for work
lm_q4f <- lm(happy ~ clove + work, data = mba)
summary(lm_q4f)
```

- The coefficient for `clove` in the regression model is `r round(summary(lm_q4f)$coef["clove", "Estimate"], 2)`, which is somewhat different from the manually computed average of `r round(mean(mean_differences), 2)`.
- The two approaches do not produce the same result because they make different assumptions and weight the data differently.
- The manual calculation gives equal weight to each `work` level regardless of how many observations it contains, whereas the regression model effectively gives more weight to `work` levels with more observations when estimating the `clove` effect.
- Additionally, the regression treats `work` as a continuous numeric predictor (*assuming a linear relationship with `happy`*), while the manual approach treats each `work` level as a separate category with no assumed relationship between them.

### Q5

#### q5a

```{r}
data(sat, package = "faraway")
plot(
  total ~ salary,
  data = sat,
  xlab = "Teachers' Salaries",
  ylab = "SAT Scores"
)
lm_q5a <- lm(total ~ salary, data = sat)
summary(lm_q5a)
abline(a = lm_q5a)
```

The plot and regression line show a negative relationship between teacher salaries and SAT scores — as teacher salaries increase, SAT scores tend to decrease. However, it would be surprising if this were a causal effect. It does not make intuitive sense that paying teachers more would cause students to perform worse. This suggests the presence of a lurking or confounding variable that explains the apparent relationship.

#### q5b

Yes, the relationship is statistically significant. The coefficient for salary is `r round(summary(lm_q5a)$coef["salary", "Estimate"], 2)` with a p-value of `r round(summary(lm_q5a)$coef["salary", "Pr(>|t|)"], 4)`, which is below the 0.05 significance level. However, statistical significance does not imply causation — a significant relationship can still be driven entirely by confounding.

#### q5c

```{r}
with(
  data = sat,
  expr = symbols(
    x = salary,
    y = total,
    circles = takers,
    inches = 1 / 5,
    bg = rgb(0.1, 0.1, 0.1, 1 / 4)
  )
)
```


The plot reveals an important pattern: states with lower teacher salaries tend to have smaller circles (*fewer test takers*) and higher SAT scores, while states with higher teacher salaries tend to have larger circles (*more test takers*) and lower SAT scores. This suggests that the proportion of test takers may be a confounding variable — in states where only a small, self-selected group of high-achieving students take the SAT, scores are naturally higher, regardless of teacher salaries.

#### q5d

```{r}
lm_q5d <- lm(total ~ takers, data = sat)
summary(lm_q5d)
```

The model shows a strong, statistically significant negative relationship between the proportion of test takers and SAT scores (coefficient = `r round(summary(lm_q5d)$coef["takers", "Estimate"], 2)`, p-value = `r format(summary(lm_q5d)$coef["takers", "Pr(>|t|)"], digits = 3)`). The $R^2$ of `r round(summary(lm_q5d)$r.squared, 4)` indicates that the proportion of takers explains a large portion of the variation in SAT scores.

It is reasonable to assume this relationship is at least partly causal. In states where only a small proportion of students take the SAT, these tend to be the most academically able students who are self-selecting into the test to gain admission to competitive colleges. As the proportion of takers increases, less academically prepared students are drawn into taking the test, which naturally lowers the average score. This is a form of selection bias — the composition of the test-taking pool changes as participation increases.

#### q5e

```{r}
with(data = sat, expr = cor(x = salary, y = takers))
```

The correlation between teacher salaries and the proportion of test takers is `r round(with(sat, cor(salary, takers)), 4)`, indicating a moderately strong positive relationship. This means that states with higher teacher salaries also tend to have a greater proportion of students taking the SAT. However, it is unlikely that one directly causes the other. Rather, both variables are likely driven by broader underlying factors such as state wealth, education policy, or geographic region.

#### q5f

```{r}
lm_q5f <- lm(total ~ salary + takers, data = sat)
summary(lm_q5f)
```

After adjusting for the proportion of test takers, the coefficient for teacher salary reverses sign — it is now positive (`r round(summary(lm_q5f)$coef["salary", "Estimate"], 2)`) and statistically significant (p = `r round(summary(lm_q5f)$coef["salary", "Pr(>|t|)"], 4)`). This means that once we account for differences in test participation rates, higher teacher salaries are associated with higher SAT scores, which is the intuitively expected direction.

- Confounding: The proportion of takers is associated with both teacher salary (as shown in (e)) and SAT scores (as shown in (d)), and its omission in (a) distorted the estimated effect of salary on scores.
- Simpson's paradox: The direction of the relationship between salary and SAT scores reverses — from negative in the unadjusted model in (b) to positive in the adjusted model here — once we control for the confounding variable takers.

### Q6

#### q6a

```{r}
data(newhamp, package = "faraway")
lm_q6a <- lm(pObama ~ votesys + Dean, data = newhamp)
summary(lm_q6a)
```

#### q6b

```{r}

```