---
title: "Linear Model with R: Chapter 7"
author: Han Chenyue
editor: source
format: html
---

```{r, echo = FALSE, message = FALSE}
library(corrplot)

knitr::opts_chunk$set(
  # echo = FALSE,
  digits = 4,
  scipen = 2
)
```

# Chapter 6

## Exercises

### Q1

#### q1a

```{r}
data(sat, package = "faraway")
colnames(sat)
```

```{r}
lm_q1 <- lm(total ~ expend + salary + ratio + takers, data = sat)
summary(lm_q1)
```

```{r}
plot(
  x = fitted(lm_q1),
  y = residuals(lm_q1),
  xlab = "Fitted",
  ylab = "Residuals"
)
abline(h = 0)
```

```{r}
plot(
  x = fitted(lm_q1),
  y = sqrt(abs(residuals(lm_q1))),
  xlab = "Fitted",
  ylab = "Root Absolute Residuals"
)
```

```{r}
plot(x = lm_q1, which = 1)
```

```{r}
plot(x = lm_q1, which = 3)
```

The points appear evenly spread, with no clear pattern suggesting unequal variance or non-linearity.

#### q1b

```{r}
qqnorm(y = residuals((lm_q1)), ylab = "Residuals", main = "")
qqline(residuals(lm_q1))
```

```{r}
plot(x = lm_q1, which = 2)
```

The Q-Q plot indicates that the data are approximately normally distributed.

#### q1c

```{r}
hatv <- hatvalues(lm_q1)
sum(hatv)
```

```{r}
us_states <- row.names(x = sat)
faraway::halfnorm(x = hatv, labs = us_states, ylab = "Leverages")
```

```{r}
names(hatv[hatv > 2 * length(coef(lm_q1)) / nrow(x = sat)])
```

Visually, `California` and `Utah` exhibit high leverage that could substantially affect the model fit. Evaluating this against the standard $2p/n$ threshold (cutoff = `r 2 * length(coef(lm_q1)) / nrow(sat)`), we find that `New Jersey` and `Connecticut` also exceed the limit. A closer look at these four states is warranted to determine whether their leverage translates into a substantive effect on the regression estimates.

#### q1d

```{r}
plot(x = lm_q1, which = 2)
```

There are no outliers seen in the Q-Q plot.

```{r}
range(rstudent(lm_q1))
```

Since the largest residual in magnitude is `r round(abs(range(rstudent(lm_q1))[1]), 2)`, which is less than $3.49$, none of the points in the dataset will be flagged as formal statistical outliers using the Bonferroni correction at the $5\%$ level.

#### q1e

```{r}
faraway::halfnorm(
  x = cooks.distance(lm_q1),
  nlab = 3,
  labs = rownames(x = sat),
  ylab = "Cook's Distances"
)
```

`Utah` might seem like an influential point at first glance, but its Cook's distance of `r max(round(cooks.distance(lm_q1), 2))` is less than one, indicating it does not disproportionately impact the model.

#### q1f

```{r}
par(mfrow = c(2, 2))
termplot(model = lm_q1, partial.resid = TRUE)
par(mfrow = c(1, 1))
```

The relationship exhibits no significant deviations from linearity

### Q2
