---
title: "Linear Model with R: Chapter 5"
author: Han Chenyue
editor: source
format: html
---

```{r, echo = FALSE, message = FALSE}
library(tidyverse)

knitr::opts_chunk$set(
  # echo = FALSE,
  digits = 4,
  scipen = 2
)
```

# Chapter 5

## Exercises

### Q1

#### q1a

```{r}
data(prostate, package = "faraway")
lm_q1a <- lm(lpsa ~ ., data = prostate)
summary(lm_q1a)
```

```{r}
new_patient_a <- data.frame(
  lcavol = 1.44692,
  lweight = 3.62301,
  age = 65.00000,
  lbph = 0.30010,
  svi = 0.00000,
  lcp = -0.79851,
  gleason = 7.00000,
  pgg45 = 15.00000
)

predict(object = lm_q1a, newdata = new_patient_a, interval = "prediction")
```

Manual calculations for sanity check:

To compute the CI manually, we will need:

$$
\hat{y}_0 \pm t^{(\alpha / 2)}_{n-p} \hat{\sigma} \sqrt{1 + x_0^T (X^T X)^{-1} x_0}
$$

$$
\hat{\sigma} = \sqrt{\frac{RSS}{n - p}} = \sqrt{\frac{\hat{e}^T \hat{e}}{n - p}} = \sqrt{\frac{\sum_{i=1}^n \hat{e}_i^2}{n - p}}
$$

```{r}
y <- prostate$lpsa
y

X <- model.matrix(lm_q1a)
head(X)

new_patient_manual_calculations <- data.frame(
  intercept = 1,
  lcavol = 1.44692,
  lweight = 3.62301,
  age = 65.00000,
  lbph = 0.30010,
  svi = 0.00000,
  lcp = -0.79851,
  gleason = 7.00000,
  pgg45 = 15.00000
)

new_patient_manual_calculations

beta_hat <- solve(t(X) %*% X) %*% t(X) %*% y
x0 <- t(as.matrix(new_patient_manual_calculations))
epsilon <- y - X %*% beta_hat
rss <- t(epsilon) %*% epsilon
n <- nrow(X)
p <- ncol(X)
sigma_hat <- sqrt(rss / (n - p))
```

```{r}
y0 <- t(x0) %*% beta_hat
y0
```

```{r}
near(
  y0,
  predict(
    object = lm_q1a,
    newdata = new_patient_manual_calculations,
    interval = "prediction"
  )[1, "fit"]
)
```

```{r}
t_critical <- qt(0.975, df = n - p)
t_critical

lwr <- y0 -
  t_critical * sigma_hat * sqrt(1 + t(x0) %*% solve(t(X) %*% X) %*% x0)
lwr
near(
  lwr,
  predict(
    object = lm_q1a,
    newdata = new_patient_manual_calculations,
    interval = "prediction"
  )[1, "lwr"]
)

upr <- y0 +
  t_critical * sigma_hat * sqrt(1 + t(x0) %*% solve(t(X) %*% X) %*% x0)
upr
near(
  upr,
  predict(
    object = lm_q1a,
    newdata = new_patient_manual_calculations,
    interval = "prediction"
  )[1, "upr"]
)
```

#### q1b

```{r}
new_patient_b <- data.frame(
  lcavol = 1.44692,
  lweight = 3.62301,
  age = 20.00000,
  lbph = 0.30010,
  svi = 0.00000,
  lcp = -0.79851,
  gleason = 7.00000,
  pgg45 = 15.00000
)

predict(object = lm_q1a, newdata = new_patient_b, interval = "prediction")
```

The youngest patient in the original dataset is aged forty-one, whereas the current patient is less than half of that age. Therefore, the prediction involves substantial extrapolation beyond the range of the observed data, making it less reliable and resulting in a wider prediction interval. Furthermore, the prediction interval is wider because it accounts for the variability of a future observation through the inclusion of the error term $\epsilon$, in addition to the uncertainty in the estimation of $\beta$.

#### q1c

```{r}
summary(lm_q1a)
```

Only `lcavol`, `lweight`, and `svi` are significant at the $5\%$ level.

```{r}
lm_q1c <- lm(lpsa ~ lcavol + lweight + svi, data = prostate)
summary(lm_q1c)
```

```{r}
old_model_for_patient_a <- predict(
  object = lm_q1a,
  newdata = new_patient_a,
  interval = "prediction"
)
new_model_for_patient_a <- predict(
  object = lm_q1c,
  newdata = new_patient_a,
  interval = "prediction"
)
```

```{r}
cat("Result for patient A (Old Model): ", old_model_for_patient_a, "\n")
cat("Result for patient A (New Model): ", new_model_for_patient_a)
```

The predicted values are very similar.

```{r}
old_model_for_patient_b <- predict(
  object = lm_q1a,
  newdata = new_patient_b,
  interval = "prediction"
)
new_model_for_patient_b <- predict(
  object = lm_q1c,
  newdata = new_patient_b,
  interval = "prediction"
)
```

```{r}
cat("Result for patient B (Old Model): ", old_model_for_patient_b, "\n")
cat("Result for patient B (New Model): ", new_model_for_patient_b)
```

In this case, the results differ more noticeably, particularly with respect to the prediction interval.

Based on these examples alone, we cannot determine which model provides better predictive performance. The reduced model, however, has the practical advantage that it requires only three predictors to produce a prediction. If obtaining the full set of predictors is difficult or costly, this simpler model may therefore be preferable.

### Q2

#### q2a

```{r}
data(teengamb, package = "faraway")
```

```{r}
lm_q2a <- lm(gamble ~ sex + status + income + verbal, data = teengamb)
summary(lm_q2a)
```

```{r}
apply(subset(teengamb, sex == 0), 2, mean)
male_with_average_stats_data <- data.frame(
  sex = 0,
  status = 52,
  income = 4.976071,
  verbal = 6.821429,
  gamble = 29.775000
)
predict(
  object = lm_q2a,
  newdata = male_with_average_stats_data,
  interval = "prediction"
)
```

The 95% prediction interval is \[-16.83, 76.38\], though in practical terms, the lower bound would be truncated at zero.

Standard linear regression assumes that our response variable `gamble` can be any number on the continuous number line, extending to infinity in both the positive and negative directions. The model doesn't "know" that our data represents money.

Furthermore, our Residual Standard Error is 22.69. This is a relatively large amount of error compared to the predicted value (29.775). Because the error is so large, subtracting it from our predicted value pushes the bottom of the interval well below zero.

#### q2b

```{r}
aggregate(teengamb[, -1], list(sex = teengamb$sex), max)
```

```{r}
male_with_maximal_stats_data <- data.frame(
  sex = 0,
  status = 75,
  income = 15,
  verbal = 10
)
```

```{r}
predict(
  object = lm_q2a,
  newdata = male_with_maximal_stats_data,
  interval = "prediction"
)
```

This CI is wider. In linear regression, our predictions are most accurate and have the tightest confidence/prediction intervals at the centre of mass of our data. As we move away from the average towards the extreme edges of our data, the model becomes less certain (the bowtie effect).

There are also fewer data points at the extremes, as there are usually many people in a dataset with "average" incomes and scores but very few who have the absolute maximum income, maximum status, and maximum verbal score all at the same time. Because the model has less data representing these extreme individuals, its predictions carry more uncertainty, forcing the interval to be wider.

#### q2c

```{r}
lm_q2c <- lm(sqrt(gamble) ~ sex + status + income + verbal, data = teengamb)
summary(lm_q2c)
```

```{r}
# Squaring the prediction is needed to reverse the square root on the response. It also solves the problem of having negative lower bound, as the interval cannot mathematically drop below zero anymore.
predict(
  object = lm_q2c,
  newdata = male_with_average_stats_data,
  interval = "prediction"
)^2
```

#### q2d

```{r}
female_with_custom_stats_data <- data.frame(
  sex = 1,
  status = 20,
  income = 1,
  verbal = 10
)
predict(
  object = lm_q2c,
  newdata = female_with_custom_stats_data,
  interval = "prediction"
)^2
```

`lwr` $>$ `upr`, which is nonsensical.

Without squaring it, the results are:

```{r}
predict(
  object = lm_q2c,
  newdata = female_with_custom_stats_data,
  interval = "prediction"
)
```

-   The fatal flaw is that the model predicted the square root of this girl's gambling spending to be `r round(predict(object = lm_q2c, newdata=female_with_custom_stats_data, interval="prediction")[1, "fit"], 2)`, with a lower bound of `r round(predict(object = lm_q2c, newdata=female_with_custom_stats_data, interval="prediction")[1, "lwr"], 2)`.
-   Mathematically, the principal square root of a real number cannot be negative. The model has made an impossible prediction.
-   Because a negative times a negative is a positive, squaring her large negative lower bound (`r round(predict(object = lm_q2c, newdata=female_with_custom_stats_data, interval="prediction")[1, "lwr"], 2)`) turns it into a large positive number (`r round(predict(object = lm_q2c, newdata=female_with_custom_stats_data, interval="prediction")[1, "lwr"]^2, 2)`).
-   A monotone mathematical function is one that always goes in the same direction (*always up or always down*). Squaring is not a monotone function across all real numbers, because it reverses the order of negative values.
-   Therefore, squaring these negative bounds distorted and flipped the interval, showing that this model performs poorly (*extrapolates badly*) for this specific demographic.

### Q3

#### q3a

```{r}
data(snail, package = "faraway")
head(snail)
```

```{r}
round(xtabs(water ~ temp + humid, data = snail) / 4, 1)
```

No, the specific temperature and humidity isn't listed in the table and we cannot interpolate with just a raw table.

#### q3b

```{r}
lm_q3b <- lm(water ~ temp + humid, data = snail)
summary(lm_q3b)
```

```{r}
predict(
  object = lm_q3b,
  newdata = data.frame(temp = 25, humid = 60),
  interval = "prediction"
)
```

Predicted water content: `r round(predict(object = lm_q3b, newdata = data.frame(temp = 25, humid = 60), interval = "prediction")[1, "fit"], 1)`%

#### q3c

```{r}
predict(
  object = lm_q3b,
  newdata = data.frame(temp = 30, humid = 75),
  interval = "prediction"
)
```

Predicted water content: `r round(predict(object = lm_q3b, newdata = data.frame(temp = 30, humid = 75), interval = "prediction")[1, "fit"], 1)`%

Predicted water content from part (a): `r round(xtabs(water ~ temp + humid, data = snail)[2, 2] / 4, 1)`%

Comparing these two predictions highlights a classic statistical tradeoff. The model-based prediction is likely a better, more stable estimate because it leverages the entire dataset to find an overall trend. However, its main drawback is the risk of bias; the model assumes the relationship between temperature, humidity, and water is perfectly linear. If that biological assumption is incorrect, the prediction will be systematically off.

Conversely, the raw average from part (a) makes no assumptions about linear relationships, meaning it is not dependent on a choice of model (*no model bias*). However, because it only uses the data from the 4 specific snails tested at those exact conditions, the small sample size makes the estimate highly susceptible to random chance and natural variation (*high variance*).

#### q3d

The intercept of `r round(coef(lm_q3b)[1], 1)`% represents the model's predicted water content when both `temp` and `humid` are equal to zero.

This answer is **not unique**. Because the multiple regression model creates a 3D linear plane, thus there are infinite combinations of predictors that will result in the exact same number. Any combination of temperature and humidity where their mathematical effects cancel each other out (i.e., where $-0.183 \times \text{temp} + 0.473 \times \text{humid} = 0$) will also yield this predicted response.

Furthermore, this is **not a reasonable prediction**. Setting the temperature to 0°C and humidity to 0% is an extreme example of **extrapolation** (*predicting far outside the range of our collected data*). Biologically, these conditions are completely abnormal and would likely be lethal to the snails.

#### q3e

Given the linear regression equation:

$$
\text{Water Content} = \beta_0 + \beta_1 \times \text{Temperature} + \beta_2 \times \text{Humidity}
$$

We can plug in our known values:

$$
80 = 52.611 - 0.183 \times 25 + 0.473 \times \text{Humidity}
$$

Next, we rearrange the equation to isolate the missing variable (`Humidity`):

$$
\text{Humidity} = \frac{80 - 52.611 - (-0.183 \times 25)}{0.473}
$$

Solving this gives a resulting humidity of `r round((80 - coef(lm_q3b)[1] - coef(lm_q3b)[2] * 25) / coef(lm_q3b)[3], 1)`%.

### Q4

#### q4a

```{r}
data(mdeaths, package = "faraway")
head(mdeaths)
```

```{r}
plot(mdeaths, ylab = "No. of Deaths")
```

During the winter season.

#### q4b

```{r}
time_series_with_current_and_lagged_13_values <- embed(
  x = mdeaths,
  dimension = 14
)
head(time_series_with_current_and_lagged_13_values)
colnames(time_series_with_current_and_lagged_13_values) <- c(
  "y",
  paste0("lag", 1:13)
)
time_series_with_current_and_lagged_13_values <- data.frame(
  time_series_with_current_and_lagged_13_values
)
lm_q4b <- lm(
  y ~ lag1 + lag12 + lag13,
  data = time_series_with_current_and_lagged_13_values
)
summary(lm_q4b)
```

Only `lag12`, the one-year lag, is statistically significant.

#### q4c

```{r}
# Getting data from the last row
time_series_with_current_and_lagged_13_values[
  nrow(time_series_with_current_and_lagged_13_values),
]
```

Given `r time_series_with_current_and_lagged_13_values[nrow(time_series_with_current_and_lagged_13_values), ]`

-   `lag1` equals `r time_series_with_current_and_lagged_13_values[nrow(time_series_with_current_and_lagged_13_values), ]["y"]`

-   `lag12` equals `r time_series_with_current_and_lagged_13_values[nrow(time_series_with_current_and_lagged_13_values), ]["lag11"]`

-   `lag13` equals `r time_series_with_current_and_lagged_13_values[nrow(time_series_with_current_and_lagged_13_values), ]["lag12"]`

```{r}
predict(
  object = lm_q4b,
  newdata = data.frame(lag1 = 1341, lag12 = 2263, lag13 = 1812),
  interval = "prediction"
)
```

#### q4d

-   `lag1` equals `r predict(object = lm_q4b, newdata = data.frame(lag1 = 1341, lag12 = 2263, lag13 = 1812), interval = "prediction")[, "fit"]`

-   `lag12` equals `r time_series_with_current_and_lagged_13_values[nrow(time_series_with_current_and_lagged_13_values), ]["lag10"]`

-   `lag13` equals `r time_series_with_current_and_lagged_13_values[nrow(time_series_with_current_and_lagged_13_values), ]["lag11"]`

```{r}
predict(
  object = lm_q4b,
  newdata = data.frame(lag1 = 1879.5994643, lag12 = 1820, lag13 = 2263),
  interval = "prediction"
)
```

#### q4e

```{r}
length(x = fitted(object = lm_q4b))
```

```{r}
length(x = mdeaths)
```

The original dataset contains `r length(x = mdeaths)` observations, but our model only produces `r length(x = fitted(object = lm_q4b))` fitted values. This discrepancy occurs because the autoregressive model relies on lagged variables (*data from previous months*). Consequently, the earliest observations in the dataset cannot be predicted because the required lag data falls before the start of our recording period.

To accurately plot the data, we must align the arrays by subsetting the observed values to match the timeframe of our `r length(x = fitted(object = lm_q4b))` fitted values `mdeaths[14:72]`.

```{r}
# 72 total - 59 fitted = 13 skipped at the beginning.
# Therefore, we plot using indices 14 through 72.
plot(
  fitted(lm_q4b),
  tail(x = mdeaths, n = length(x = fitted(object = lm_q4b))),
  xlab = "Fitted Values",
  ylab = "Observed Values"
)
abline(0, 1)
```

-   Based on the plot, the prediction accuracy will not be the same for all months of the year.

-   We can observe the presence of **heteroscedasticity** in the plot: the variability (*vertical spread*) of the points increases as the total number of deaths increases. Because the `mdeaths` dataset is highly seasonal, certain months of the year, typically winter months, experience drastically higher death tolls. Therefore, the model's predictions for those high-count months will have a wider variance and lower absolute accuracy compared to the predictions for low-count months.

-   It is important to understand *why* this occurs. An increased number of deaths does not mean the model is inherently failing. Rather, winter months introduce greater real-world volatility (*such as unpredictable flu outbreaks*), making the baseline harder to predict. Furthermore, statistical errors are often proportional; a standard percentage error yields a much larger absolute numerical difference when applied to a high winter death toll compared to a low summer baseline. This proportional error manifests as the wider visual spread seen on the graph.

### Q5

#### q5a

```{r}
data(fat, package = "faraway")
head(fat)
```

To determine if the smaller model is justifiable, we can fit both models and compare them using an $F$-test.

```{r}
lm_q5a_i <- lm(brozek ~ age + weight + height + abdom, data = fat)
lm_q5a_ii <- lm(
  brozek ~ age +
    weight +
    height +
    neck +
    chest +
    abdom +
    hip +
    thigh +
    knee +
    ankle +
    biceps +
    forearm +
    wrist,
  data = fat
)
anova(lm_q5a_i, lm_q5a_ii)
```

When comparing a reduced model to a full model, we are testing whether the 9 extra coefficients ($\beta$) in the full model are exactly zero.

-   $H_0: \beta_{neck} = \beta_{chest} = \dots = \beta_{wrist} = 0$ (*The extra variables do not significantly improve the model; the smaller model is sufficient*).
-   $H_a:$ At least one of these 9 coefficients is $\neq 0$ (*The extra variables provide a significant improvement; the full model is better*).

The ANOVA test yields an $F$-statistic of 2.934 and a highly significant p-value of 0.00256 ($p < 0.01$). Because $p < \alpha$, we reject $H_0$. Statistically, the extra variables provide a significant improvement, meaning the full model is mathematically superior.

Another way to evaluate the models is by comparing their residual standard errors, which measure prediction uncertainty.

```{r}
# Comparing their residual standard errors, which measure prediction uncertainty
c(
  Smaller_Model = summary(lm_q5a_i)$sigma,
  Full_Model = summary(lm_q5a_ii)$sigma
)
```

The full model has a slightly lower prediction error (`r round(summary(lm_q5a_ii)$sigma, 2)`) compared to the smaller model (`r round(summary(lm_q5a_i)$sigma, 2)`).

From a strict statistical perspective, the smaller model is not justifiable because the $F$-test explicitly rejects it in favor of the larger model.

#### q5b

```{r}
x <- data.frame(t(apply(X = fat, MARGIN = 2, FUN = median)))
x
```

```{r}
predict(object = lm_q5a_i, newdata = x, interval = "prediction")
predict(object = lm_q5a_ii, newdata = x, interval = "prediction")
```

No.

#### q5c

```{r}
fat[25:50, ]
```

-   Subject 39 is unusually heavy.
-   Subject 42 is unusually short.

#### q5d

```{r}
lm_q5d_i <- lm(
  brozek ~ age + weight + height + abdom,
  data = fat,
  subset = -c(39, 42)
)
lm_q5d_ii <- lm(
  brozek ~ age +
    weight +
    height +
    neck +
    chest +
    abdom +
    hip +
    thigh +
    knee +
    ankle +
    biceps +
    forearm +
    wrist,
  data = fat,
  subset = -c(39, 42)
)

predict(object = lm_q5d_i, newdata = x, interval = "prediction")
predict(object = lm_q5d_ii, newdata = x, interval = "prediction")
```

No, removing the extreme outliers made almost no difference. Because our prediction is based on the median (*the exact center*) of the dataset, it is highly stable. Outliers strongly pull on the extremes/tails of a regression model, but they have very little leverage over predictions made directly in the middle of the data spread.

### Q6

#### q6a

```{r}
data(fat, package = "faraway")
lm_q6a <- lm(
  brozek ~ age +
    weight +
    height +
    neck +
    chest +
    abdom +
    hip +
    thigh +
    knee +
    ankle +
    biceps +
    forearm +
    wrist,
  data = fat
)
prediction_q6a <- predict(
  object = lm_q6a,
  newdata = fat,
  interval = "prediction"
)
```

$$
RMSE = \sqrt{\frac{1}{m} \sum_{j = 1}^m (\hat{y}_j - \tilde{y}_j)^2}
$$

```{r}
sqrt(mean((prediction_q6a[, "fit"] - fat$brozek)^2))
```

#### q6b

```{r}
mean(
  (fat$brozek > prediction_q6a[, "lwr"]) &
    (fat$brozek < prediction_q6a[, "upr"])
)
```

#### q6c

```{r}
validation_set <- fat[1:50, ]
training_set <- fat[51:nrow(fat), ]
```

```{r}
lm_q6c <- lm(
  brozek ~ age +
    weight +
    height +
    neck +
    chest +
    abdom +
    hip +
    thigh +
    knee +
    ankle +
    biceps +
    forearm +
    wrist,
  data = training_set
)
prediction_q6c <- predict(
  object = lm_q6c,
  newdata = training_set,
  interval = "prediction"
)
```

Allocating 50 out of 252 observations (`r round(50/252*100, 1)`%) to the validation set is a standard and appropriate sample size, leaving ample data (202 observations) to train a mathematically stable model. However, the *method* of selecting the first 50 consecutive rows is a dangerous practice especially if the dataset happens to be sorted by any variable, a sequential split guarantees that the validation set will have a fundamentally different underlying distribution than the training set. To ensure both sets are representative of the same population, the observations should always be partitioned using random sampling.

#### q6d

```{r}
sqrt(mean((prediction_q6c[, "fit"] - training_set$brozek)^2))
```

The RMSE on the training set (`r round(sqrt(mean((prediction_q6c[, "fit"] - training_set$brozek)^2)), 3)`) is slightly lower than the RMSE on the full dataset (`r round(sqrt(mean((prediction_q6a[, "fit"] - fat$brozek)^2)), 3)`). This decrease is mathematically expected because we are measuring **in-sample error**. Ordinary Least Squares (OLS) regression explicitly calculates coefficients designed to minimize the squared residuals for the exact data it is trained on. Furthermore, because the model retains all 13 predictors but now has fewer observations to fit (202 instead of 252), it gains slightly more flexibility to "*hug*" the training data, inherently driving the training error down.

```{r}
mean(
  (training_set$brozek > prediction_q6c[, "lwr"]) &
    (training_set$brozek < prediction_q6c[, "upr"])
)
```

The fraction of prediction intervals containing the true response on the training set is (`r round(mean((training_set$brozek > prediction_q6c[, "lwr"]) & (training_set$brozek < prediction_q6c[, "upr"])) * 100, 1)`%), which is remarkably close to the (`r round(mean((fat$brozek > prediction_q6a[, "lwr"]) & (fat$brozek < prediction_q6a[, "upr"])) * 100, 1)`%) coverage we saw on the full dataset. Both of these empirical coverage rates are slightly conservative (*running a bit higher than the nominal 95% target*), but they prove that the model's behavior remains stable and consistent when trained on this subset. 

#### q6e

```{r}
prediction_q6e <- predict(
  object = lm_q6c,
  newdata = validation_set,
  interval = "prediction"
)
sqrt(mean((prediction_q6e[, "fit"] - validation_set$brozek)^2))
```

The RMSE on the unseen validation set (`r round(sqrt(mean((prediction_q6e[, "fit"] - validation_set$brozek)^2)), 3)`) is noticeably higher than the RMSE on the training set (`r round(sqrt(mean((prediction_q6c[, "fit"] - training_set$brozek)^2)), 3)`). This is an expected statistical phenomenon. Regression models are mathematically optimised to minimise the error on their specific training data. When the model is forced to generalise to brand-new, unseen data, the error naturally increases. In other words, the training RMSE represents an overly optimistic view of the model's accuracy, whereas the validation RMSE provides a much more realistic estimate of real-world performance.

```{r}
mean(
  (validation_set$brozek > prediction_q6e[, "lwr"]) &
    (validation_set$brozek < prediction_q6e[, "upr"])
)
```

The coverage probability on the validation set actually rose slightly to `r round(mean((validation_set$brozek > prediction_q6e[, "lwr"]) & (validation_set$brozek < prediction_q6e[, "upr"])) * 100, 1)`%. While we typically expect coverage to drop slightly on unseen data, this anomaly is easily explained by the small sample size. Because the validation set contains exactly `r nrow(validation_set)` observations, a single observation represents a 2% swing. A `r round(mean((validation_set$brozek > prediction_q6e[, "lwr"]) & (validation_set$brozek < prediction_q6e[, "upr"])) * 100, 1)`% coverage means exactly 49 out of 50 intervals contained the true response. If just one more observation had fallen outside the interval, the coverage would be 96%, perfectly mirroring the training set. Therefore, the coverage remains highly stable despite the small sample size variance.

### Q7

#### q7a

```{r}
# Defining a general RSME function
rmse_function <- function(a, b) {
  sqrt(mean((a - b)^2))
}
```

```{r}
data(fat, package = "faraway")
lm_q7a <- lm(
  siri ~ age +
    weight +
    height +
    neck +
    chest +
    abdom +
    hip +
    thigh +
    knee +
    ankle +
    biceps +
    forearm +
    wrist,
  data = fat
)
prediction_q7a <- predict(
  object = lm_q7a,
  newdata = fat,
  interval = "prediction"
)
```

```{r}
rmse_function(fat$siri, prediction_q7a[, "fit"])
```

The RMSE of the model on the training data is `r round(rmse_function(fat$siri, prediction_q7a[, "fit"]), 3)`.

```{r}
mean(
  (fat$siri > prediction_q7a[, "lwr"] & (fat$siri < prediction_q7a[, "upr"]))
)
```

The fraction of prediction intervals containing the observed response is `r round(mean((fat$siri > prediction_q7a[, "lwr"]) & (prediction_q7a[, "upr"] > fat$siri)) * 100, 1)`%. This empirical coverage slightly exceeds the nominal 95% target, indicating a well-calibrated, slightly conservative model when evaluated in-sample.

#### q7b

```{r}
data(wfat, package = "faraway")
prediction_q7b <- predict(
  object = lm_q7a,
  newdata = wfat,
  interval = "prediction"
)
rmse_function(wfat$siri, prediction_q7b[, "fit"])
```

When applying the men's model to the women's dataset, the RMSE jumps drastically to `r round(rmse_function(wfat$siri, prediction_q7b[, "fit"]), 3)` (*nearly double the in-sample error*). This massive increase indicates that a model trained on male physiology is fundamentally inaccurate at predicting female body fat percentages.

```{r}
mean(
  (wfat$siri > prediction_q7b[, "lwr"]) & (wfat$siri < prediction_q7b[, "upr"])
)
```

Interestingly, despite the poor accuracy (*high RMSE*), the coverage fraction remains excellent at `r round(mean((wfat$siri > prediction_q7b[, "lwr"]) & (prediction_q7b[, "upr"] > wfat$siri)) * 100, 1)`%. The model successfully captured the true values within its intervals, but as we will see in part (e), it only achieved this by making the intervals incredibly wide.

#### q7c

```{r}
plot(
  x = prediction_q7b[, "fit"],
  y = wfat$siri,
  xlab = "Predicted Body Fat From Model",
  ylab = "Actual Body Fat From Data"
)
abline(a = 0, b = 1)
```

#### q7d

```{r}
# Mean difference for Women (Out-of-sample)
mean(prediction_q7b[, "fit"] - wfat$siri)
```

The mean difference for the women's data is `r round(abs(mean(wfat$siri - prediction_q7b[, "fit"])), 1)`%. This quantifies our visual observation from part (c): on average, a woman has about `r round(abs(mean(wfat$siri - prediction_q7b[, "fit"])), 1)`% more body fat than a man with the exact same physical dimensions.

```{r}
# Mean difference for Men (In-sample)
mean(prediction_q7a[, "fit"] - fat$siri)
```

Conversely, the mean difference for the men is functionally zero as the sum (*and therefore the mean*) of the residuals on the training data is always exactly zero.

#### q7e

```{r}
# Mean width of prediction intervals for Men
mean(prediction_q7a[, "upr"] - prediction_q7a[, "lwr"])
```

```{r}
# Mean width of prediction intervals for Women
mean(prediction_q7b[, "upr"] - prediction_q7b[, "lwr"])
```

The average prediction interval width for men is `r round(mean(prediction_q7a[, "upr"] - prediction_q7a[, "lwr"]), 1)`, while for women it expands massively to `r round(mean(prediction_q7b[, "upr"] - prediction_q7b[, "lwr"]), 1)`.

This perfectly explains the paradox from part (b). The mathematical formula for a prediction interval incorporates the distance of the new observation from the centroid of the training data. Because women's physical measurements are structurally different from the male averages the model was trained on, the algorithm applies a mathematical "penalty" for predicting in unfamiliar territory, causing the intervals to balloon outward. This immense uncertainty is the only reason the coverage probability survived.

#### q7f

This exercise demonstrates both quantitative and qualitative extrapolation. Women have a distinct range of physical measurements (*quantitative*) and a fundamentally different biological physiology (*qualitative*). Consequently, the linear relationships learned from men fail when applied to a different demographic.

Also, relying solely on training data creates a false sense of security. Testing the model on a completely distinct population (*external validation*) successfully revealed that our model's predictive power is strictly limited to men and cannot be generalised to the broader public.